Drought frequency predicts life history strategies in African \emph{Heliophila}

### 2018-11-30 manuscript/manuscript.tex
updates, edits, spell checked

\documentclass[man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Drought frequency predicts plant life history strategies},
            pdfauthor={J. Grey Monroe, Brian Gill, Kathryn G. Turner, \& John K. McKay},
            pdfkeywords={drought adaptation, life history evolution, remote sensing,
phylogeography, herbaria records},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Drought frequency predicts plant life history strategies}
    \author{J. Grey Monroe\textsuperscript{1,2}, Brian Gill\textsuperscript{3},
Kathryn G. Turner\textsuperscript{4}, \& John K.
McKay\textsuperscript{2}}
    \date{}

\shorttitle{Drought and life history}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO 80523, USA\\\textsuperscript{2} College of Agriculture, Colorado State University, Fort Collins, CO 80523, USA\\\textsuperscript{3} Institute for Environment and Society, Brown University, Providence, RI 02912, USA\\\textsuperscript{4} Biology Department, Pennsylvania State University, State College, PA 16802, USA}
\keywords{drought adaptation, life history evolution, remote sensing, phylogeography, herbaria records}
\usepackage{csquotes}
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

Correspondence concerning this article should be addressed to J. Grey
Monroe, 307 University Ave, Fort Collins, CO 80523. E-mail:
\href{mailto:monroejg@colostate.edu}{\nolinkurl{monroejg@colostate.edu}}}

\abstract{
Explaining variation in life history strategies is a long-standing goal
of evolutionary biology. For plants, annual and perennial life histories
are thought to reflect adaptation to environments that differ in the
frequency of environmental stress such as drought. Here we test this
hypothesis in \textit{Heliophila} (Brassicaceae), a diverse genus of
flowering plants native to Africa by integrating 2192 herbaria
occurrence records with 34 years of satellite-based drought detection.
Consistent with predictions from classic life history theory, we find
that perennial \textit{Heliophila} species occur in environments where
droughts are significantly less frequent compared to annuals. These
associations are predictive while controlling for phylogeny, lending
support to the hypothesis that drought related natural selection has
influenced the distributions of these strategies. Additionally, the
difference in drought frequency between annual and perennial species
distributions is greatest during the summer and fall, which also appears
to be when annuals are in the seed phase of their life cycle based on
collection dates of annual species. Together, these finding provide
empirical support for classic hypotheses about the drivers of life
history strategy in plants - that perennials out compete annuals in
environments with less frequent drought and that annuals are adpated to
enviroments with more frequent drought by escaping drought prone seasons
as seeds. 
}

@@

Understanding the causes and consequences of life history variation is a
longstanding goal of ecology and evolutionary biology (Cole, 1954). In
plants, life histories are especially diverse, with herbaceous species
that complete their life cycle in a number of weeks to trees that live
for thousands of years (Brown, 1996). Along this continuum an important
division exists, distinguishing annuals which complete their seed to
seed life cycle within a single calendar year from perennials which can
persist over multiple years. Annual plants flower once, set seed,
senesce, and then die, spending at least some portion of the year as a
seed, where they are relatively protected from environmental stress. In
contrast, perennial plants can continue vegetative growth after
reproduction and must survive conditions experienced during all seasons.
These represent fundamentally different life history strategies, but the
ecological factors that explain their evolution and distributions remain
empirically uresolved (Friedman and Rubin, 2015).\\
Classical theory predict shorter life spans in environments where adult
mortality is high (Charnov and Schaffer, 1973; Franco and Silvertown,
1996; Stearns, 1992). In plants, this has been extended to the
hypothessis that annuality is adaptive when it allows plants to escape
drought (Schaffer and Gadgil, 1975). Lack of water is perhaps the
greatest threat to survival during vegetatitive or reproductive growth
and annuals can remain dormant (and protected as a seed) during drought.
Thus, environments with greater seasonal drought frequency may select
for annual life histories that complete reproduction prior to drought
prone seasons. Conversely, environments with less frequent drought may
select for perennial species, which may benefit from multiple bouts of
reproduction and competitive advantage by preventing recruitment of
annual species (Corbin and D'Antonio, 2004). These predictions have been
supported by the association of annualilty with arid environments in
\emph{Oryza perennis} (Morishima et al., 1984) and \emph{Oenothera}
(Evans et al., 2005). Additionally, annual and perennial species of
\emph{Nemesia} were qualitatively associated with winter rather and
summer rainfall environments respectively (Datson et al., 2008) and
annual species of \emph{Scorzoneroides} were associated with
environments classified as unpredictable (Cruz-Mazo et al., 2009).
However, whether the history frequency of drought events indeed predicts
the distributions annual or perennial life history strategies has yet to
be tested in a phylogenetic context with suitable sample sizes of
distrubution and climate data.\\
Here we combine a long-term global dataset of satellite detected drought
events with metadata from natural history collections to test these
classic hypotheses about the evolution of life history strategies within
the African endemic mustard genus, \emph{Heliophila} L. (Brassicaceae).
If annuallity is an adaptive strategy allowing plants to escape drought
prone seasons, then drought frequency should predict the distribution of
life history strategies across landscapes, and annual species should be
more commonly associated with drought prone regions than perennial
species. Furthermore, if annual species have adapted to escape drought
prone seasons, observations of growing annual species (i.e.~occurring in
forms other than seed) should be rare during drought prone seasons.
Phylogenetic relatedness can have significant non-random effects on
species distributions and life history traits (Barrett et al., 1996),
and therefore we assessed the relationship between life history
distribution and drought frequency in a phylogenetically controlled
background.

@@

All analyses were peformed using R. All data and the source code to
produce this manuscript are available at
\url{https://github.com/greymonroe/heliophila}. Softare used is listed
in the supplement.

@@

Thus, VHI measurements are standardized according to conditions
historically observed at each locations. These measurements have been
validated and generally used for evaluating drought risk and predicting
crop yields in agriculture (e.g., Rojas et al., 2011; Kogan et al.,
2016). But they also present a new tool to study seasonal patterns in
the frequency of drought across environments and to test hypotheses
about the effect of drought on ecological and evolutionary processes
(Kerr and Ostrovsky, 2003). As such, the VHI has been applied recently
to study drought related ecology of natural species and proven useful
for predicting infraspecific variation in drought tolerance traits and
genes (Dittberner et al., 2018; Mojica et al., 2016; Monroe et al.,
2018). Here, we accessed VHI data at \(16km^2\) resolution from 1981 to
2015 (\url{https://www.star.nesdis.noaa.gov/smcd/emb/vci/VH/vh_ftp.php})
to characterize the seasonal drought frequencies experienced by annual
and perennial \emph{Heliophila} species.

@@

\emph{Heliophila} is a genus of flowering plants endemic to the southern
portion of Africa including the Cape Floristic and Succulent Karoo
Regions. These are among the most botanically diverse environments on
Earth and the estimated \textasciitilde{}50 \emph{Heliophila} species
occurring there are considered to make up the most diverse genus of the
family Brassicaceae (Mandáková et al., 2012; Mummenhoff et al., 2005).
This genus includes both perennial and annual species and this change in
life history strategy has likely arisen multiple independent times
(Appel and Al-Shehbaz, 1997; Mummenhoff et al., 2005). Furthermore, the
fine scale climatic heterogeneity of Southern Africa is ideal for
studying the distribution of traits in relation to environmental
parameters (Sayre et al., 2013). We used life histories reported by
Mummenhoff et al. (2005), grouping species with annual or perennial life
histories. Perenniality was defined based any form of perennial life
history (e.g., herbs, shrubs, mixed, etc). We also repeated these
analyses without \emph{H. subulata} and \emph{H. minima} which have
mixed life histories and found similar results. Because the original
authors didn't report the nature of the mixed trait (Mummenhoff et al.,
2005), (ie. plasticity vs.~genetic variation), we classified these
species here as perennial since they have the capacity to maintain
vegetative growth after reproduction at least to some capacity.

\hypertarget{heliophila-occurance-records}{%
\subsubsection{Heliophila occurance
records}\label{heliophila-occurance-records}}

Botanists have collected and maintained over 350 million botanical
specimens worldwide over the past 300 years. Herbarium specimens and
their associated metadata have been used since the 1960s to study
species' geographical distributions (reviewed by Willis et al. (2017)
and Lang et al. (2018)). And as they become digitized (Soltis, 2017),
these collections have been used to study relationships between trait
distributions, geography, and climate (Davis et al., 2015; Stropp et
al., 2016; Václavı'k et al., 2017; Wolf et al., 2016). To characterize
the diributions of annual and perennial \emph{Heliophila} species, all
records for the genus \emph{Heliophila} were downloaded from the Global
Biodiversity Information Facility (gbif.org) on July 21, 2018 (GBIF,
2018).

@@

To characterize drought regimens across the distrubtions of annual and
perennial species of \emph{Heliohpila}, we calculated drought during
different seasons at the location of observations for \emph{Heliophila}
records using the VHI. Specifically, we created global maps of the
frequencies of observing drought conditions (VHI\textless{}40, NOAA)
during the winter (quarter surrounding winter solstice), spring (quarter
surrounding spring equinox), summer (quarter surrounding summer
solstice) and fall (quarter surrounding fall equinox) from 1981 to 2015.
From these maps, the drought frequency during the winter, spring,
summer, and fall were extracted for the locations of all GBIF records.

\hypertarget{filtering-of-occurance-records}{%
\subsubsection{Filtering of occurance
records}\label{filtering-of-occurance-records}}

@@

\begin{itemize}
\tightlist
\item
  records for species with reported life history\\
\item
  records with geospatial data\\
\item
  records without known geospatial coodinate issues (i.e., coordinates
  reported are those of herbarium)\\
\item
  records from collection sites classified as land pixels\\
\item
  records from Africa (to exclude locations of cultivation)
\item
  records without duplicates (i.e., identical species, location,
  collection date)
\end{itemize}

@@

To evaluate the hypothesis that annual and perennial life history
strategies reflect adaptations to alternative drought regimes, we tested
the corresponding prediction that the observed distributions of annual
and perennial \emph{Heliophila} species would be significantly
associated with historic drought frequency. First, we compared the
frequency of drought during the winter, spring, summer, and fall between
raw occurance records of annual and perennial species by t-tests. To
account for variation in the number of occurance records per species, we
next calculated the mean drought frequency during the winter, spring,
summer and fall for each species. The relationships between species mean
values of drought frequency during each season and life habitat (annual
or perennial) were tested using Firth's penalized-likelihood logistic
regressions. Because demographic histories caused by ancestry can
confound trait - environment associations, we then tested for the
relationships between drought frequeny and life history while
controlling for relationships between species using phylogenetic
logistic regressions.

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../figures/maps_boxplots} \caption{Locations of (a) annual and (b) perennial Heliophila.
Drought frequency during the (c) winter, (d) spring, (e) summer and (f)
fall. (g) Drought frequencies during each season observed at the
collection sites of \emph{Heliophila} records.}\label{fig:mapsboxplots}
\end{figure}

Out of 8670 \emph{Heliophila} GBIF recrods, 6634 were for species with
reported life history (Mummenhoff et al., 2005), 3653 had geospatial
data, 3627 did not have geospatial issues, 3460 were located on pixels
classified as land having drought measurements, 3457 were located in
Africa, 3162 were not duplicated. After all filtering steps, 2192
records for 42 species (Figure \ref{fig:phylogeny}, Table
\ref{tab:speciesmeanstable}) passed for further analyses. The number of
samples varied between species, with a mean of 52.19 samples per
species. Heliophila rigidiuscula had the most records, 201, and
Heliophila cornellsbergia the fewest, 2 (Table
\ref{tab:speciesmeanstable}).

There were clear visual differences between the distributions of the 960
annual and the 1232 perennial \emph{Heliophila} observation records (see
Figure \ref{fig:speciesmaps} for maps of individual species). While
annual species were generally found in the western regions of South
Africa and Namibia, primarily in the Cape Floristic Region and Succulent
Karoo (Figure \ref{fig:mapsboxplots}a), the occurances of perennials
extended to the east coast of South Africa (Figure
\ref{fig:mapsboxplots}b).

The frequency of drought varied considerably across the ranges of
\emph{Heliophila} species (Figure \ref{fig:mapsboxplots}c-f). This
heterogeneity is expected, given that this is one of the most
climatically diverse regions of the Earth (Sayre et al., 2013). It is
worth noting the east to west cline in drought frequency observed during
the summer, which distinguishes the high drought frequency of the
Kalahari Sands and Namid Destert phyotogeographic regions from the low
drought frequency of the Drakensberg Mountains and Coastal Zambesian
phytogeographic regions. In the Cape phytogeographic region there was
finer scale heterogeneity in drought frequency during the summer.

Theory predicts that annuality should be adaptive in places where
stresses such as drought are more common. Conversely, perenniality
should be adaptive in places where such stresses are less frequent. We
found that the frequency of drought was significantly higher at the
locations of occurance records for annual species. In terms of raw
observation records (Figure \ref{fig:mapsboxplots}g), the frequency of
drought was signficantly higher at the location of annuals during the
winter (t = 10.65, p = 0.00), spring (t = 10.73, p = 0.00), summer (t =
12.67, p = 0.00), and fall (t = 15.26, p = 0.00). Because raw occurance
records do not account for variation in the number of records per
species (Table \ref{tab:speciesmeanstable}), we also tested whether mean
drought frequency values of each species were significantly different
between annuals and perennials by a Firths-penalized logistic
regression. We found that the mean drought frequencies were signficantly
higher (\(\alpha = 0.05\)) in annual species during the spring, summer,
and fall (Table \ref{tab:modelstable}, Figure \ref{fig:lineplots}a). We
further tested whether annual species are found in places where droughts
occur more frequently while controlling for the phylogentic relatedness
of \emph{Heliophila} species. This is important, because environmental
differences in species distributions can be confounded with demographic
history caused by acestry. We found that while controlling for
phylogeny, the mean drought frequencies were signficantly higher
(\(\alpha = 0.05\)) in annual species during the spring, summer, and
fall (Table \ref{tab:modelstable}, Figure \ref{fig:lineplots}a). These
findings indicate that common acestry alone does not explain differences
the drought frequencies experienced in the environemnts of annual and
perennial \emph{Heliophila}.

@@

\begin{figure}
\centering
\includegraphics{../figures/line_and_dates.pdf}
\caption{\label{fig:lineplots}(a) Comparison (mean +- SE) of drought frequency across
seasons measured at the GBIF records of annual and perennial species of
Heliophila. (b) Collection dates of GBIF records of annual and perennial
species of Heliophila.}
\end{figure}

The preceding results indicated that annual speices are found in
environments where droughts are signifacantly more frequent, especially
in the summer and fall. Classic life history theory hypothesizes that
annuality reflects adaptation to such environments because it allows
species to escape stressful conditions. If this is the case, we would
expect that annuals spend the drought prone seasons of summer and fall
as seeds. To test this hypothesis, we compared the dates of occurance
records between annual and perennial \emph{Heliophila} species. The
distributions reveal a considerable difference in the timing of
observation of these two life histories. In comparison to perennials,
which appear to be collected throughout the year, annuals are almost
exclusively observed during the winter and spring (Figure
\ref{fig:lineplots}b). The differences between the distrubution of
collection dates were signicant by all tests (ks.test D = 0.25, p = 0;
bartlett.test K2 = 503.18, p = 0.00) This is consistent with a model of
life history in which annual species flower in the spring, set seed,
senesce, and die before the summer. Thus, these annual species are
likely to remain dormant during the summer and fall, when drought is the
strongest predictor of the distrubutions of annual and perennial life
histories (Figure \ref{fig:lineplots}a).

@@

>>>>

\documentclass[man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Drought frequency predicts life history strategies in African Heliophila},
            pdfauthor={J. Grey Monroe, Brian Gill, Kathryn G. Turner, \& John K. McKay},
            pdfkeywords={drought adaptation, life history evolution, remote sensing,
phylogeography, herbaria records},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Drought frequency predicts life history strategies in African
\emph{Heliophila}}
    \author{J. Grey Monroe\textsuperscript{1,2}, Brian Gill\textsuperscript{3},
Kathryn G. Turner\textsuperscript{4}, \& John K.
McKay\textsuperscript{2}}
    \date{}

\shorttitle{Drought and life history}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO 80521, USA\\\textsuperscript{2} College of Agriculture, Colorado State University, Fort Collins, CO 80521, USA\\\textsuperscript{3} Institute for Environment and Society, Brown University, Providence, RI 02912, USA\\\textsuperscript{4} Biology Department, Pennsylvania State University, State College, PA 16802, USA}
\keywords{drought adaptation, life history evolution, remote sensing, phylogeography, herbaria records}
\usepackage{csquotes}
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

Correspondence concerning this article should be addressed to J. Grey
Monroe, 307 University Ave, Fort Collins, CO 80521. E-mail:
\href{mailto:monroejg@colostate.edu}{\nolinkurl{monroejg@colostate.edu}}}

\abstract{
Explaining variation in life history strategies is a long-standing goal
of evolutionary biology. For plants, annual and perennial life histories
are thought to reflect adaptation to environments that differ in the
frequency of environmental stress such as drought. Here we test this
hypothesis in \textit{Heliophila} (Brassicaceae), a diverse genus of
flowering plants native to Africa, by integrating 34 years of
satellite-based drought measurements with 2192 herbaria occurrence
records. Consistent with predictions from classic life history theory,
we find that perennial \textit{Heliophila} species occur in environments
where droughts are significantly less frequent compared to annuals.
These associations are predictive while controlling for phylogeny,
lending support to the hypothesis that drought related natural selection
has influenced the distributions of these strategies. Additionally, the
difference in drought frequency between annual and perennial species
distributions is greatest during the summer and fall, which also appears
to be when annuals are in the seed phase of their life cycle based on
collection dates of annual species. Together, these finding provide
empirical support for classic hypotheses about the drivers of life
history strategy in plants - that perennials out compete annuals in
environments with less frequent drought and that annuals are adapted to
environments with more frequent drought by escaping drought prone
seasons as seeds. 
}

@@

Understanding the causes and consequences of life history variation is a
longstanding goal of ecology and evolutionary biology (Cole, 1954). In
plants, life histories are especially diverse, with herbaceous species
that complete their life cycle in a number of weeks to trees that live
for thousands of years (Brown, 1996). Along this continuum an important
division exists, distinguishing annuals which complete their seed to
seed life cycle within a single calendar year from perennials which can
persist over multiple years. Annual plants flower once, set seed,
senesce, and then die, spending at least some portion of the year as a
seed, where they are relatively protected from environmental stress. In
contrast, perennial plants can continue vegetative growth after
reproduction and must survive conditions experienced during all seasons.
These represent fundamentally different life history strategies, but the
ecological factors that explain their evolution and distributions remain
empirically uresolved (Friedman and Rubin, 2015).\\
Classical theory predict shorter life spans in environments where adult
mortality is high (Charnov and Schaffer, 1973; Franco and Silvertown,
1996; Stearns, 1992). In plants, this has been extended to the
hypothesis that annuality is adaptive when it allows plants to escape
drought (Schaffer and Gadgil, 1975). Lack of water is perhaps the
greatest threat to survival during vegetative or reproductive growth and
annuals can remain dormant (and protected as a seed) during drought.
Thus, environments with greater seasonal drought frequency may select
for annual life histories that complete reproduction prior to drought
prone seasons. Conversely, environments with less frequent drought may
select for perennial species, which may benefit from multiple bouts of
reproduction and competitive advantage by preventing recruitment of
annual species (Corbin and D'Antonio, 2004). These predictions have been
supported by the observation of annuals in arid environments in
\emph{Oryza perennis} (Morishima et al., 1984) and \emph{Oenothera}
(Evans et al., 2005). Additionally, annual and perennial species of
\emph{Nemesia} were qualitatively associated with winter rather and
summer rainfall environments respectively (Datson et al., 2008) and
annual species of \emph{Scorzoneroides} were associated with
environments classified as unpredictable (Cruz-Mazo et al., 2009).
However, whether the history frequency of drought events indeed predicts
the distributions annual or perennial life history strategies has yet to
be tested in a phylogenetic context with suitable sample sizes of
distribution and climate data.\\
Here we combine a long-term global dataset of satellite detected drought
events with metadata from natural history collections to test these
classic hypotheses about the evolution of life history strategies within
the African endemic mustard genus, \emph{Heliophila} L. (Brassicaceae).
If annuality is an adaptive strategy allowing plants to escape drought
prone seasons, then drought frequency should predict the distribution of
life history strategies across landscapes, and annual species should be
more commonly associated with drought prone regions than perennial
species. Furthermore, if annual species have adapted to escape drought
prone seasons, observations of growing annual species (i.e.~occurring in
forms other than seed) should be rare during drought prone seasons.
Phylogenetic relatedness can have significant non-random effects on
species distributions and life history traits (Barrett et al., 1996),
and therefore we assessed the relationship between life history
distribution and drought frequency in a phylogenetically controlled
background.

@@

All analyses were performed using R. All data and the source code to
produce this manuscript are available at
\url{https://github.com/greymonroe/heliophila}. Software used is listed
in the supplement.

@@

Thus, VHI measurements are standardized according to conditions
historically observed at each locations. These measurements have been
validated and generally used for evaluating drought risk and predicting
crop yields in agriculture (e.g., Rojas et al., 2011; Kogan et al.,
2016). But they also present a new tool to study seasonal patterns in
the frequency of drought across environments and to test hypotheses
about the effect of drought on ecological and evolutionary processes
(Kerr and Ostrovsky, 2003). As such, the VHI has been applied recently
to study drought related ecology of natural species and proven useful
for predicting intraspecific variation in drought tolerance traits and
genes (Dittberner et al., 2018; Mojica et al., 2016; John G Monroe et
al., 2018). Here, we accessed VHI data at \(16km^2\) resolution from
1981 to 2015
(\url{https://www.star.nesdis.noaa.gov/smcd/emb/vci/VH/vh_ftp.php}) to
characterize the seasonal drought frequencies experienced by annual and
perennial \emph{Heliophila} species.

@@

\emph{Heliophila} is a genus of flowering plants endemic to the southern
portion of Africa including the Cape Floristic and Succulent Karoo
Regions. These are among the most botanically diverse environments on
Earth and the estimated \textasciitilde{}50 \emph{Heliophila} species
occurring there are considered to make up the most diverse genus of the
family Brassicaceae (Mandáková et al., 2012; Mummenhoff et al., 2005).
This genus includes both perennial and annual species and this change in
life history strategy has likely arisen multiple independent times
(Appel and Al-Shehbaz, 1997; Mummenhoff et al., 2005). Furthermore, the
fine scale climatic heterogeneity of Southern Africa is ideal for
studying the distribution of traits in relation to environmental
parameters (Sayre et al., 2013). We used life histories reported by
Mummenhoff et al. (2005), grouping species with annual or perennial life
histories. Perenniality was defined based any form of perennial life
history (e.g., herbs, shrubs, mixed, etc). Because the nature of species
reported with mixed traits were unknown (i.e.~plasticity vs.~genetic
variation), we classified these species here as perennial since they can
maintain vegetative growth after reproduction at least to some capacity.

\hypertarget{heliophila-occurrence-records}{%
\subsubsection{Heliophila occurrence
records}\label{heliophila-occurrence-records}}

Botanists have collected and maintained over 350 million botanical
specimens worldwide over the past 300 years. Herbarium specimens and
their associated metadata have been used since the 1960s to study
species' geographical distributions (reviewed by Willis et al. (2017)
and Lang et al. (2018)). And as they become digitized (Soltis, 2017),
these collections have been used to study relationships between trait
distributions, geography, and climate (Davis et al., 2015; Stropp et
al., 2016; Václavı'k et al., 2017; Wolf et al., 2016). To characterize
the distributions of annual and perennial \emph{Heliophila} species, all
records for the genus \emph{Heliophila} were downloaded from the Global
Biodiversity Information Facility (gbif.org) on July 21, 2018 (GBIF,
2018).

@@

To characterize drought regimens across the distributions of annual and
perennial species of \emph{Heliophila}, we calculated drought during
different seasons at the location of observations for \emph{Heliophila}
records using the VHI. Specifically, we created global maps of the
frequencies of observing drought conditions (VHI\textless{}40, NOAA)
during the winter (quarter surrounding winter solstice), spring (quarter
surrounding spring equinox), summer (quarter surrounding summer
solstice) and fall (quarter surrounding fall equinox) from 1981 to 2015.
From these maps, the drought frequency during the winter, spring,
summer, and fall were extracted for the locations of all GBIF records.

\hypertarget{filtering-of-occurrence-records}{%
\subsubsection{Filtering of occurrence
records}\label{filtering-of-occurrence-records}}

@@

\begin{itemize}
\tightlist
\item
  records for species with reported life history\\
\item
  records with geospatial data\\
\item
  records without known geospatial coordinate issues (i.e., coordinates
  reported are those of herbarium)\\
\item
  records from collection sites classified as land pixels\\
\item
  records from Africa (to exclude locations of cultivation)
\item
  records without duplicates (i.e., identical species, location,
  collection date)
\end{itemize}

@@

To evaluate the hypothesis that annual and perennial life history
strategies reflect adaptations to alternative drought regimes, we tested
the corresponding prediction that the observed distributions of annual
and perennial \emph{Heliophila} species would be significantly
associated with historic drought frequency. First, we compared the
frequency of drought during the winter, spring, summer, and fall between
raw occurrence records of annual and perennial species by t-tests. To
account for variation in the number of occurrence records per species,
we next calculated the mean drought frequency during the winter, spring,
summer and fall for each species. The relationships between species mean
values of drought frequency during each season and life habitat (annual
or perennial) were tested using Firth's penalized-likelihood logistic
regressions. Because demographic histories caused by ancestry can
confound trait - environment associations, we then tested for the
relationships between drought frequency and life history while
controlling for relationships between species using phylogenetic
logistic regressions.

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../figures/maps_boxplots} \caption{Locations of occurrence records of (a) annual and (b)
perennial \emph{Heliophila.} Drought frequency during the (c) winter,
(d) spring, (e) summer and (f) fall measured using the VHI. (g) Drought
frequencies during each season at the observation locations of annual
and perennial \emph{Heliophila} (t tests, ** = p \textless{} 0.01).}\label{fig:mapsboxplots}
\end{figure}

Out of 8670 \emph{Heliophila} GBIF records, 6634 were for species with
reported life history (Mummenhoff et al., 2005), 2856 had geospatial
data, 2833 did not have geospatial issues, 2684 were located on pixels
classified as land having drought measurements, 2543 were located in
Africa, 2192 were not duplicated. Thus, after all filtering steps, 2192
records for 42 species (Figure \ref{fig:phylogeny}, Table
\ref{tab:speciesmeanstable}) passed for further analyses. The number of
samples varied between species, with a mean of 52.19 samples per
species. Heliophila rigidiuscula had the most records, 201, and
Heliophila cornellsbergia the fewest, 2 (Table
\ref{tab:speciesmeanstable}).

There were clear visual differences between the distributions of the 960
annual and the 1232 perennial \emph{Heliophila} observation records (see
Figure \ref{fig:speciesmaps} for maps of individual species). While
annual species were generally found in the western regions of South
Africa and Namibia, primarily in the Cape Floristic Region and Succulent
Karoo (Figure \ref{fig:mapsboxplots}a), the occurrence of perennials
extended to the east coast of South Africa (Figure
\ref{fig:mapsboxplots}b).

The frequency of drought varied considerably across the ranges of
\emph{Heliophila} species (Figure \ref{fig:mapsboxplots}c-f). This
heterogeneity is expected, given that this is one of the most
climatically diverse regions of the Earth (Sayre et al., 2013). It is
worth noting the east to west cline in drought frequency observed during
the summer, which distinguishes the high drought frequency of the
Kalahari Sands and Namid Desert phytogeographic regions from the low
drought frequency of the Drakensberg Mountains and Coastal Zambesian
phytogeographic regions. In the Cape phytogeographic region there was
finer scale heterogeneity in drought frequency during the summer.

Theory predicts that annuality should be adaptive in places where
stresses such as drought are more common. Conversely, perenniality
should be adaptive in places where such stresses are less frequent. We
found that the frequency of drought was significantly higher at the
locations of occurrence records for annual species. In terms of raw
observation records (Figure \ref{fig:mapsboxplots}g), the frequency of
drought was significantly higher at the location of annuals during the
winter (t = 10.65, p = 0.00), spring (t = 10.73, p = 0.00), summer (t =
12.67, p = 0.00), and fall (t = 15.26, p = 0.00). Because raw occurrence
records do not account for variation in the number of records per
species (Table \ref{tab:speciesmeanstable}), we also tested whether mean
drought frequency values of each species were significantly different
between annuals and perennials by a Firths-penalized logistic
regression. We found that the mean drought frequencies were
significantly higher (\(\alpha = 0.05\)) in annual species during the
spring, summer, and fall (Table \ref{tab:modelstable}, Figure
\ref{fig:lineplots}a). We further tested whether annual species are
found in places where droughts occur more frequently while controlling
for the phylogentic relatedness of \emph{Heliophila} species. This is
important, because environmental differences in species distributions
can be confounded with demographic history caused by ancestry. We found
that while controlling for phylogeny, the mean drought frequencies were
significantly higher (\(\alpha = 0.05\)) in annual species during the
spring, summer, and fall (Table \ref{tab:modelstable}, Figure
\ref{fig:lineplots}a). These findings indicate that common acestry alone
does not explain differences the drought frequencies experienced in the
environments of annual and perennial \emph{Heliophila}.

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../figures/line_and_dates} \caption{(a) Comparison (mean +- SE) of drought frequency across
seasons measured at the GBIF records of annual and perennial species of
Heliophila. (phylogenetic logistic regression, * = p \textless{} 0.05,
** = p \textless{} 0.01) (b) Collection dates of GBIF records of annual
and perennial species of Heliophila.}\label{fig:lineplots}
\end{figure}

The preceding results indicated that annual species are found in
environments where droughts are significantly more frequent, especially
in the summer and fall. Classic life history theory hypothesizes that
annuality reflects adaptation to such environments because it allows
species to escape stressful conditions. If this is the case, we would
expect that annuals spend the drought prone seasons of summer and fall
as seeds. To test this hypothesis, we compared the dates of occurrence
records between annual and perennial \emph{Heliophila} species. The
distributions reveal a considerable difference in the timing of
observation of these two life histories. In comparison to perennials,
which appear to be collected throughout the year, annuals are almost
exclusively observed during the winter and spring (Figure
\ref{fig:lineplots}b). The differences between the distribution of
collection dates were significant by all tests (ks.test D = 0.25, p = 0;
bartlett.test K2 = 503.18, p = 0.00) This is consistent with a model of
life history in which annual species flower in the spring, set seed,
senesce, and die before the summer. Thus, these annual species are
likely to remain dormant during the summer and fall, when drought is the
strongest predictor of the distributions of annual and perennial life
histories (Figure \ref{fig:lineplots}a).

@@

\hypertarget{author-contributions}{%
\section{Author contributions}\label{author-contributions}}
